---
title: "Hands-on Exercise 10: Modelling, Visualising and Analysing Network Data with R"
author: "Nguyen Bao Thu Phuong"
date: "26 March 2025" 
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# Getting Started

The below code chunk loads the relevant package into R environment.

```{r}
pacman::p_load(lubridate, ggthemes, reactable,
reactablefmtr, gt, gtExtras, tidyverse)
```

# Importing Microsoft Access database

## The data set

A personal database in Microsoft Access mdb format called *Coffee Chain* will be used for this exercise.

## Importing database into R

The code chunk below uses `odbcConnectAccess()` of [**RODBC**](https://cran.r-project.org/web/packages/RODBC/index.html) package to import the database query table into R.

```{r}
#| eval: false
library(RODBC)
con <- odbcConnectAccess2007('data/Coffee Chain.mdb')
coffeechain <- sqlFetch(con, 'CoffeeChain Query')
write_rds(coffeechain, "data/CoffeeChain.rds")
odbcClose(con)
```

## Data Preparation

The code chunk below is used to import *CoffeeChain.rds* into R.

```{r}
coffeechain <- read_rds("data/CoffeeChain.rds")
```

The code chunk below aggregates Sales and Budgeted Sales at the Product level.

```{r}
product <- coffeechain %>%
  group_by(`Product`) %>%
  summarise(`target` = sum(`Budget Sales`),
            `current` = sum(`Sales`)) %>%
  ungroup()
```

## Bullet chart in ggplot2

The code chunk below plots the bullet charts using ggplot2 functions.

```{r}
ggplot(product, aes(Product, current)) + 
  geom_col(aes(Product, max(target) * 1.01),
           fill="grey85", width=0.85) +
  geom_col(aes(Product, target * 0.75),
           fill="grey60", width=0.85) +
  geom_col(aes(Product, target * 0.5),
           fill="grey50", width=0.85) +
  geom_col(aes(Product, current), 
           width=0.35,
           fill = "black") + 
  geom_errorbar(aes(y = target,
                    x = Product, 
                    ymin = target,
                    ymax= target), 
                width = .4,
                colour = "red",
                size = 1) +
  coord_flip()
```

# Plotting sparklines using ggplot2

## Preparing the data

```{r}
sales_report <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  mutate(Month = month(Date)) %>%
  group_by(Month, Product) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup() %>%
  select(Month, Product, Sales)
```

The code chunk below computes the minimum, maximum and end othe the month sales.

```{r}
mins <- group_by(sales_report, Product) %>% 
  slice(which.min(Sales))
maxs <- group_by(sales_report, Product) %>% 
  slice(which.max(Sales))
ends <- group_by(sales_report, Product) %>% 
  filter(Month == max(Month))
```

The code chunk below is computes the 25 and 75 quantiles.

```{r}
quarts <- sales_report %>%
  group_by(Product) %>%
  summarise(quart1 = quantile(Sales, 
                              0.25),
            quart2 = quantile(Sales, 
                              0.75)) %>%
  right_join(sales_report)
```

## sparklines in ggplot2

```{r}
ggplot(sales_report, aes(x=Month, y=Sales)) + 
  facet_grid(Product ~ ., scales = "free_y") + 
  geom_ribbon(data = quarts, aes(ymin = quart1, max = quart2), 
              fill = 'grey90') +
  geom_line(size=0.3) +
  geom_point(data = mins, col = 'red') +
  geom_point(data = maxs, col = 'blue') +
  geom_text(data = mins, aes(label = Sales), vjust = -1) +
  geom_text(data = maxs, aes(label = Sales), vjust = 2.5) +
  geom_text(data = ends, aes(label = Sales), hjust = 0, nudge_x = 0.5) +
  geom_text(data = ends, aes(label = Product), hjust = 0, nudge_x = 1.0) +
  expand_limits(x = max(sales_report$Month) + 
                  (0.25 * (max(sales_report$Month) - min(sales_report$Month)))) +
  scale_x_continuous(breaks = seq(1, 12, 1)) +
  scale_y_continuous(expand = c(0.1, 0)) +
  theme_tufte(base_size = 3, base_family = "Helvetica") +
  theme(axis.title=element_blank(), axis.text.y = element_blank(), 
        axis.ticks = element_blank(), strip.text = element_blank())
```

# Static Information Dashboard Design: gt and gtExtras methods

This section covers creating static information dashboards using the **gt** and **gtExtras** packages. Before getting started, it is highly recommended to explore their official webpages to gain an overview of their purposes and available functions. While a deep understanding is not required at this stage, familiarizing yourself with their capabilities will be beneficial.

## Plotting a simple bullet chart

The below code chunk plots a bullet chart report using functions of gt and gtExtras packages.

```{r}
product %>%
  gt::gt() %>%
  gt_plt_bullet(column = current, 
              target = target, 
              width = 60,
              palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# sparklines: gtExtras method

The code chunk below prepares the data for gtExtras functions.

```{r}
report <- coffeechain %>%
  mutate(Year = year(Date)) %>%
  filter(Year == "2013") %>%
  mutate (Month = month(Date, 
                        label = TRUE, 
                        abbr = TRUE)) %>%
  group_by(Product, Month) %>%
  summarise(Sales = sum(Sales)) %>%
  ungroup()
```

It is important to note that most **gtExtras** functions require a data.frame with list columns. To meet this requirement, the code chunk below converts the report data.frame into the appropriate format.

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

## Plotting Coffechain Sales report

```{r}
report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop") %>%
   gt() %>%
   gt_plt_sparkline('Monthly Sales',
                    same_limit = FALSE)
```

## Adding statistics

the below code chunk calculate summary statistics

```{r}
report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            ) %>%
  gt() %>%
  fmt_number(columns = 4,
    decimals = 2)
```

## Combining the data.frame

The below code chunk add the relevant statistics to the table.

```{r}
spark <- report %>%
  group_by(Product) %>%
  summarize('Monthly Sales' = list(Sales), 
            .groups = "drop")
```

```{r}
sales <- report %>% 
  group_by(Product) %>% 
  summarise("Min" = min(Sales, na.rm = T),
            "Max" = max(Sales, na.rm = T),
            "Average" = mean(Sales, na.rm = T)
            )
```

```{r}
sales_data = left_join(sales, spark)
```

## Plotting the updated data.table

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales',
                   same_limit = FALSE)
```

## Combining bullet chart and sparklines

Similarly, we can combining the bullet chart and sparklines using the steps below.

```{r}
bullet <- coffeechain %>%
  filter(Date >= "2013-01-01") %>%
  group_by(`Product`) %>%
  summarise(`Target` = sum(`Budget Sales`),
            `Actual` = sum(`Sales`)) %>%
  ungroup() 
```

```{r}
sales_data = sales_data %>%
  left_join(bullet)
```

```{r}
sales_data %>%
  gt() %>%
  gt_plt_sparkline('Monthly Sales') %>%
  gt_plt_bullet(column = Actual, 
                target = Target, 
                width = 28,
                palette = c("lightblue", 
                          "black")) %>%
  gt_theme_538()
```

# Interactive Information Dashboard Design: reactable and reactablefmtr methods

In this section, we explore how to create an interactive information dashboard using the **reactable** and **reactablefmtr** packages. Before getting started, it is recommended to explore their official webpages to gain an overview of their functionalities. While a deep understanding is not required, familiarizing yourself with their purpose and available functions will be helpful.

To incorporate interactive sparklines, install the **dataui** R package using the code chunk below.

```{r}
#| echo: false
remotes::install_github("timelyportfolio/dataui")
```

Next, we load the package onto R environment using the code chunk below.

```{r}
library(dataui)
```

## Plotting interactive sparklines

Similar to gtExtras, to plot an interactive sparklines using **reactablefmtr** package, we need to prepare the list field using the code chunk below.

```{r}
report <- report %>%
  group_by(Product) %>%
  summarize(`Monthly Sales` = list(Sales))
```

Next, [react_sparkline](https://kcuilla.github.io/reactablefmtr/reference/react_sparkline.html) is used plot the sparklines as shown below.

```{r}
reactable(
  report,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Changing the pagesize

The code chunk below changes defaultPageSize from 10 to 13.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(report)
    )
  )
)
```

## Adding points and labels

In the code chunk below, the `highlight_points` argument is used to display the minimum and maximum values, while the `label` argument labels the first and last values.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        labels = c("first", "last")
        )
    )
  )
)
```

## Adding reference line

The code chunk below uses `statline` argument to show the mean line.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        statline = "mean"
        )
    )
  )
)
```

## Adding bandline

Instead of adding reference line, bandline can be added using the `bandline` argument.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkline(
        report,
        highlight_points = highlight_points(
          min = "red", max = "blue"),
        line_width = 1,
        bandline = "innerquartiles",
        bandline_color = "green"
        )
    )
  )
)
```

## Changing from sparkline to sparkbar

Instead of displaying the values as sparklines, we can display them as sparkbars as shown below.

```{r}
reactable(
  report,
  defaultPageSize = 13,
  columns = list(
    Product = colDef(maxWidth = 200),
    `Monthly Sales` = colDef(
      cell = react_sparkbar(
        report,
        highlight_bars = highlight_bars(
          min = "red", max = "blue"),
        bandline = "innerquartiles",
        statline = "mean")
    )
  )
)
```
