---
title: "Hands-on Exercise 8.2: Visualising Geospatial Point Data"
author: "Nguyen Bao Thu Phuong"
date: "12 March 2025" 
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# Overview

Proportional symbol maps (also called graduated symbol maps) use symbol size to represent variations in the magnitude of a discrete, rapidly changing phenomenon, such as population counts. Similar to choropleth maps, they can be classed or unclassed:

-   Classed (Range-Graded or Graduated Symbols): Symbols are grouped into predefined size categories.

-   Unclassed (Proportional Symbols): Symbol areas are directly proportional to the mapped attribute values.

In this hands-on exercise, you will learn to create a proportional symbol map in R using the **tmap** package to visualize the number of wins at Singapore Pools’ outlets.

# Getting Started

The below code chunk use `p_load()` to load in the relevant R packages.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Geospatial Data Wrangling

## The Data

This hands-on exercise use a dataset called *SGPools_svy21.*csv.

The dataset consists of seven columns, where the XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

## Data Import and Preparation

The code chunk below uses *read_csv()* function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame.

```{r}
sgpools <- read_csv("data/aspatial/SGPools_svy21.csv")
```

After importing the data file into R, we use `list()` to examine if the data file has been imported correctly.

```{r}
list(sgpools) 
```

## Creating a sf data frame from an aspatial data frame

The code chunk below converts sgpools data frame into a simple feature data frame using `st_as_sf()` of **sf** packages.

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

Key Takeaways from the arguments above:

-   `coords` Argument: The column name for x-coordinates must be specified first, followed by the y-coordinates.

-   `crs` Argument: Requires the coordinate system in EPSG format. For example, EPSG:3414 represents Singapore’s SVY21 Projected Coordinate System. You can find EPSG codes for other countries at [epsg.io](https://epsg.io).

The basic information of the newly created sgpools_sf can be displayed as below.

```{r}
list(sgpools_sf)
```

The output shows that `sgppols_sf` is in point feature class with epsg ID = 3414. The bbox provides information of the extend of the geospatial data.

# Drawing Proportional Symbol Map

First we turn on the **view** mode of **tmap**.

```{r}
tmap_mode("view")
```

## It all started with an interactive point symbol map

The code chunks below creates an interactive point symbol map.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = 1,
           border.col = "black",
           border.lwd = 1)
```

## Lets make it proportional

Next we need to assign a numerical variable to the size visual attribute. The code chunk below assigns `Gp1Gp2Winnings` size visual attribute.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "red",
           size = "Gp1Gp2 Winnings",
           border.col = "black",
           border.lwd = 1)
```

## Lets give it a different colour

The proportional symbol map can be further improved by using the colour visual attribute. The code chunks below assignes `OUTLET_TYPE` as the colour attribute variable.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)
```

##  I have a twin brothers :)

**Tmap**’s view mode also works with faceted plots. The argument *sync* in `tm_facets()` can be used in this case to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Before ending the session, we switch **tmap**’s view back to plot mode using the code chunk below.

```{r}
tmap_mode("plot")
```
