---
title: "Hands-on Exercise 8.1: Choropleth Mapping with R"
author: "Nguyen Bao Thu Phuong"
date: "12 March 2025" 
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  message: false
  freeze: true
  warning: false
---

# Overview

Thematic mapping uses map symbols to represent specific characteristics of geographic features that are not naturally visible, such as population, temperature, crime rates, and property prices. This allows us to visualize data that would otherwise remain unseen.

On the other hand, geovisualization provides graphical representations to make places, phenomena, or processes more understandable. It leverages the power of spatial cognition associated with our eye-brain vision system, enhancing our ability to process and interpret spatial information.

In this chapter, we explore how to create accurate and meaningful choropleth maps using the **tmap** package in R.

# Import R Packages

*p_load()* of pacman package is used to install and load **sf**, **tmap** and **tidyverse** packages into R environment.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

# Import Data into R

## Data Acquisition

-   Master Plan 2014 Subzone Boundary (Web) (`MP14_SUBZONE_WEB_PL`) from <a href="https://www.data.gov.sg/" target="_blank">data.gov.sg</a>: this geospatial data is in ESRI shapefile format, which consists of the geographical boundary of Singapore at the planning subzone level. The data is based on the URA Master Plan 2014.

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 (`respopagesextod2011to2020.csv`) from <a href="https://www.singstat.gov.sg/" target="_blank">Department of Statistics, Singapore</a>: this is aspatial data in csv format. Although it does not contain any coordinates values, its `PA` and `SZ` fields can be used as unique identifiers to geocode to MP14_SUBZONE_WEB_PL shapefile.

## Import Geospatial Data into R

`MP14_SUBZONE_WEB_PL` shapefile: *st_read()* function of sf package is used to import this shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz = st_read(dsn = "data/geospatial", 
                  layer = "MP14_SUBZONE_WEB_PL")
```

We can quickly examine the content of `mpsz` using below code.

```{r}
mpsz
```

**Note:** only the first 10 records are shown, as this is the default display of R to minimize resource usage and avoid overwhelming user. To specify the number of rows to be shown, *head()* function with `n` arguments can be used instead. For example, *head(mpsz, n = 15)* shows the first 15 rows of the `mpsz` data frame .

## Import Attribute Data into R

`respopagsex2011to2020.csv` file: the aspatial data file is imported into R as a dataframe named `popdata` using *read_csv()* function of readr package as below.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

# Data Preparation

A data table with year 2020 values is prepared to facilitate plotting of the thematic map. The data table includes the following variables: `PA`, `SZ`, `YOUNG`, `ECONOMY ACTIVE`, `AGED`, `TOTAL`, `DEPENDENCY`; whereas:

-   YOUNG: age group 0 to 4 until age group 20 to 24

-   ECONOMY ACTIVE: age group 25-29 until age group 60-64

-   AGED: age group 65 and above

-   TOTAL: all age group

-   DEPENDENCY: the ratio between YOUNG and AGED against ECONOMY ACTIVE group

## Data Wrangling

The data table is prepared following below steps:

-   *filter()*, *group_by()*: filter for records in year 2020 and group by PA, SZ, AG columns.

-   *summarise()*: for each combination of PA, SZ, AG, calculates the total population (POP) by summing up the values in the `Pop` column.

-   *ungroup()*: removes the grouping, making the data available for further manipulation without any group context.

-   *pivot_wider()*: reshapes the data from long to wide format. It creates new columns for each unique value in the `AG` (age group) column, and the values in these new columns are populated with the corresponding POP values.

-   *mutate()*: calculates the assigned variable by summing the values from respective columns. For example, total population of YOUNG age group is calculated by summing the values in columns 3 to 6 and column 12.

-   *select()*: selects the final columns to include in the result `popdata2020` data frame.

```{r}
#| output: false
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup()%>%
  pivot_wider(names_from=AG, 
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[14])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:13])+
rowSums(.[15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%  
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`, 
       `ECONOMY ACTIVE`, `AGED`, 
       `TOTAL`, `DEPENDENCY`)
```

## Join the Attribute Data and Geospatial Data

Before performing the georelational join, we convert the values in `PA` and `SZ` fields to uppercase, as the original values in these fields are made up of upper and lowercase, while `SUBZONE_N` and `PLN_AREA_N` values are in uppercase. The below code converts the values to upper case and filter for records where the population in `ECONOMY ACTIVE` column is more than 0.

```{r}
popdata2020 <- popdata2020 %>%
  mutate_at(.vars = vars(PA, SZ), 
          .funs = list(toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, *left_join()* of dplyr is used to join the geographical data and attribute table using planning subzone name e.g. `SUBZONE_N` and `SZ` as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

**Note**: *left_join()* of dplyr package is used with `mpsz` simple feature data frame as the left data table to ensure that the output will be a simple features data frame.

Lastly, *write_rds()* is used to write the combined data frame into an RDS file. Ensure the respective folders are available before running the below code.

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

# Choropleth Mapping of Geospatial Data with **tmap**

Choropleth mapping represents data for geographic areas, such as countries, provinces, states, counties, or census units, using patterns or graduated colors. For example, a social scientist might use a choropleth map to show the distribution of the aged population in Singapore according to the Master Plan 2014 Subzone Boundary.

There are two approaches to creating a thematic map using the **tmap** package in R:

-   Use `qtm()` to quickly plot a basic thematic map.

-   Use various tmap elements to create a highly customizable thematic map.

## Plot a Choropleth Map quickly using *qtm()*

The easiest and quickest way to draw a choropleth map with tmap is using *qtm()*. It provides a concise default visualisation in many cases.

The below code chunk plots a cartographic standard choropleth map.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020, 
    fill = "DEPENDENCY")
```

**Things to learn**:

-   `tmap_mode()` with “plot” option is used to produce a static map. For interactive mode, “view” option should be used.

-   `fill` argument is used define the attribute to plot on the map (i.e. DEPENDENCY)

## Create a Choropleth Map using tmap’s elements

Although using `qtm()` is fast and easy, it makes aesthetics of individual layers harder to control. tmap’s drawing elements are used to draw a high quality cartographic choropleth map as shown in below figure.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "Dependency ratio") +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```

In the following sub-sections, we explore different tmap functions that can be used to plot these elements.

::: panel-tabset
### Base map (*tm_shape()*)

The basic building block of tmap is *tm_shape()*, followed by one or more layer elemments like *tm_fill()* and *tm_polygons()*.

In the below code chunk, *tm_shape()* is used to define the input data (`mpsz_pop2020`) while *tm_polygons()* is used to draw the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

### *tm_polygons()*

A choropleth map showing the geographical distribution of a selected variable by planning subzone can be drawn using *tm_polygons()* as in below code, where the coloring is based on `DEPENDENCY` values.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

**Things to learn**:

-   The default interval binning used to draw a choropleth map is called “pretty”. A detailed discussion of the data classification methods supported by tmap will be provided in sub-section 5.3.

-   The default colour scheme used is *YlOrRd* of ColorBrewer. We explore more color scheme in sub-section 5.4.

-   Missing value is shaded in grey by default.

### *tm_fill()* and *tm_border()*

*tm_polygons()* is a wrapper of *tm_fill()* and *tm_border()*. *tm_fill()* shades the polygons using the default colour scheme and *tm_borders()* adds the borders from the shapefile to the choropleth map.

The below code chunk draws a choropleth map by using *tm_fill()* only.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

We can see that the planning subzones are shaded according to their `DEPENDENCY` values, but no border is plotted.

To add the boundary of planning subzones, *tm_borders* is added as shown in below code chunk.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY") +
  tm_borders(lwd = 0.1,  alpha = 1)
```

Key arguments in *tm_borders()*:

-   *alpha*: define transparency of the line. It takes in a number between 0 (totally transparent) and 1 (not transparent). By default, the alpha value of the col is used (normally 1).

-   *col*: define border colour

-   *lwd*: define border line width. The default value is 1.

-   *lty*: define border line type. The default value is “solid”.\
:::

## Data classification methods of **tmap**

Most choropleth maps provide different data classification methods to organize a large number of observations into specific ranges or classes.

The **tmap** package offers ten data classification methods: fixed, sd, equal, pretty (default), quantile, kmeans, hclust, bclust, fisher, and jenks.

To select a data classification method, use the *style* argument in *tm_fill()* or *tm_polygons()*.

### Built-in classification methods

The below tabs shows the choropleth maps drawn using different data classification methods with the same number of classes as 5 (n=5).

::: panel-tabset
#### quantile

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "quantile") +
  tm_borders(alpha = 0.5)
```

#### equal

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "equal") +
  tm_borders(alpha = 0.5)
```

#### pretty

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "pretty") +
  tm_borders(alpha = 0.5)
```

#### jenks

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "jenks") +
  tm_borders(alpha = 0.5)
```

#### sd

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "sd") +
  tm_borders(alpha = 0.5)
```

#### hclust

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "hclust") +
  tm_borders(alpha = 0.5)
```

#### fisher

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "fisher") +
  tm_borders(alpha = 0.5)
```

#### bclust

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "bclust") +
  tm_borders(alpha = 0.5)
```

#### kmeans

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 5,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```
:::

It is clear that different data classification methods result in significantly different coloring pattern. Certain methods like *pretty*, *equal*, *sd* are more sensitive to outliers. These methods are more likely to give result with only a few areas having different colors if the data contains large outliers. Whereas methods like *quantile*, *jenks*, *kmeans* give a more evenly distributed coloring in the choropleth map.

Hence it is important to select the appropriate classification method based on the values distribution of the given dataset to avoid having misleading representation of the data in choropleth map.

The below tabs show choropleth maps drawn using the same classification method (kmeans) but with different numbers of classes (2, 6, 10, 20)

::: panel-tabset
#### n=2

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 2,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

#### n=6

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

#### n=10

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 10,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```

#### n=20

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 20,
          style = "kmeans") +
  tm_borders(alpha = 0.5)
```
:::

It is obvious that using different number of classes of the same data classification method also impact the coloring of the output map. When n=2, the output map is impacted by outliers, as only 1 area has different color. For higher number of classes, the data is distributed more evenly.

### Custom break data classification

For all built-in methods, category breaks are calculated automatically. However, we can override these defaults by specifying custom breakpoints using the *breaks* argument in *tm_fill()*. In tmap, the breakpoints must include both the minimum and maximum values. This means if you want to create n categories, you need to provide n+1 breakpoints in increasing order.

Before setting custom breakpoints, it's a good practice to examine descriptive statistics for the variable. The code snippet below calculates and displays the descriptive statistics for the `DEPENDENCY` field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

Based on the above results, we set breakpoints at 0.60, 0.70, 0.80, and 0.90. We include a minimum and maximum values, which is at 0 and 1.00 respectively. Hence our breaks is c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

Next, we plot the choropleth map using the code snippet below.

```{r}
#| warning: false
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(alpha = 0.5)
```

## Color Scheme

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the RColorBrewer package.

### ColourBrewer palette

To change the colour, we assign the preferred colour to *palette* argument of *tm_fill()* as shown in the below code chunk.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          n = 6,
          style = "quantile",
          palette = "Blues") +
  tm_borders(alpha = 0.5)
```

To reverse the colour shading, we add a “-” prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "-Greens") +
  tm_borders(alpha = 0.5)
```

The color scheme is reversed in above map, where the areas with lower `DEPENDENCY` values have darker colors.

## Map Layouts

Map layout is the arrangement of all map elements, such as the mapped objects, title, scale bar, compass, margins, and aspect ratios, to create a cohesive map. Color settings and data classification methods, like palettes and breakpoints, influence the map's appearance.

::: panel-tabset
### Map Legend

In tmap, several legend options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "jenks", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(main.title = "Distribution of Dependency Ratio by planning subzone \n(Jenks classification)",
            main.title.position = "center",
            main.title.size = 1,
            legend.height = 0.45, 
            legend.width = 0.35,
            legend.outside = FALSE,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

### Map Style

tmap allows a wide variety of layout settings to be changed using *tmap_style()*.

The code chunk below shows the *classic* style is used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "-Greens") +
  tm_borders(alpha = 0.5) +
  tmap_style("classic")
```

### Cartographic Furniture

In addition to map styling, tmap also offers options to add other map elements, such as a compass, scale bar and grid lines.

In the code snippet below, *tm_compass()*, *tm_scale_bar()*, and *tm_grid()* are used to add a compass, scale bar, and grid lines to the choropleth map.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues",
          title = "No. of persons") +
  tm_layout(main.title = "Distribution of Dependency Ratio \nby planning subzone",
            main.title.position = "center",
            main.title.size = 1.2,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar(width = 0.15) +
  tm_grid(lwd = 0.1, alpha = 0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS", 
             position = c("left", "bottom"))
```
:::

The code chunk below reset the map to default style.

```{r}
#| warning: false
tmap_style("white")
```

## Draw Small Multiple Choropleth Maps

Small multiple maps, also known as facet maps, consist of several maps arranged side by side or stacked vertically. They are useful for visualizing how spatial relationships change in relation to another variable, such as time.

In tmap, there are three ways to create small multiple maps:

-   Assign multiple values to at least one aesthetic argument.

-   Use *tm_facets()* to define a group-by variable.

-   Use *tmap_arrange()* to combine multiple stand-alone maps.

::: panel-tabset
### Assign multiple values to at least one aesthetic argument

The below code snippet creates small multiple choropleth maps by defining ncols in *tm_fill()*.

```{r}
#| warning: false
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal", 
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In this example, small multiple choropleth maps are created by assigning multiple values to at least one of the aesthetic arguments (style, palette, legend position).

```{r}
tm_shape(mpsz_pop2020)+ 
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"), 
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

### Define a group-by variable in *tm_facets()*

Multiple small choropleth maps can be created using *tm_facets().* The below code snipped creates multiple maps splitted by `REGION_N` field.

```{r}
#| warning: false
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) + 
  tm_facets(by="REGION_N", 
            free.coords=TRUE, 
            drop.shapes=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"), 
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

### Create multiple stand-alone maps with *tmap_arrange()*

In this example, multiple small choropleth maps are created from multiple stand-alone maps on the same row using *tmap_arrange()*.

```{r}
youngmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("YOUNG", 
              style = "quantile", 
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+ 
  tm_polygons("AGED", 
              style = "quantile", 
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```
:::

## Map Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth map, we can use selection funtion to map spatial objects meeting the selection criterion. The below code chunk draws the map for `REGION_N` = `CENTRAL REGION`

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY", 
          style = "quantile", 
          palette = "Blues", 
          legend.hist = TRUE, 
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45, 
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```
